Non-word repetition in Yélî Dnye

### 2020-12-15 manuscript.tex
reknit for formatting check
Note: reference castro1998 missing

\abstract{
In nonword repetition (NWR) studies, participants are presented
auditorily with an item that is phonologically legal but lexically
meaningless in the local language. NWR scores are thought to reflect
long-term phonological knowledge as well as online phonological working
memory and flexible production patterns. In this study, we report on NWR
results among children learning Yêly Dnyé, an isolate spoken in Rossel
Island, PNG, with an unusually dense phonological inventory. This study
contributes to three lines of research. First, we document that non-word
items containing typologically rare sounds are repeated without changes
less often that non-words containing more common sounds. Second, we
document rather weak effects of item length, contributing to other
research suggesting that length effects may be language-specific. Third,
we do not find strong individual variation effects in this population,
contrary to previous results documenting strong age-related effects.
Together, these data provide a unique view of online phonological
processing in a seldom-studied language, and contribute to both
typological and language acquisition research.

@@

@@

\begin{itemize}
\tightlist
\item
  double check demo -- probably start with the xlsx version that has the
  children's first names, date of birth, and date of test. I looked
  again at my local files (still haven't checked the labs' ones) and I'm
  surprised I have maternal education for more children than I have
  birth order for\ldots{} If you correct this, to generate the csv I
  also round the age, assign age.rounded to reported age (in years) when
  DOB is missing, remove first names, DOB, DOTest and age.precise before
  pushing to github
\item
  double check my frequency entries in segments.xlsx using the search
  function in \url{https://phoible.org/parameters}; take a look at
  \url{http://phoible.github.io/conventions/} in case you see something
  about double articulation (I didn't find tp or lBj)
\item
  double check also stim\_seg\_freq -- why are some items still with NA
  freq?
\item
  draft discussion
\end{itemize}

\begin{itemize}
\tightlist
\item
  add information about proportion of mispronunciations that are real
  based on other papers
\item
  add effect sizes to all analyses
\item
  ? correct treatment of words whose mispronunciation is an English
  word: mispronunciation meaning 361 two two 362 two two 879 no-pinêt
  English: no peanut 908 nodedi English: no daddy 933 no--dadi English:
  no daddy 949 petrumi English: ``true'' 950 petrumi English: ``true''
  951 petrumi English: ``true'' 1373 wee English: where 1374 wee
  English: where 1375 wee English: where 1379 wee English: where 1380
  wee English: where 1388 wee English: where 1394 wee English: where
  1396 wee English: where 1398 wee English: where 1399 wee English:
  where 1402 wat English: what 1405 wat English: what 1411 wee English:
  where
\end{itemize}

@@

Although infants begin to learn about their native language's phonology
within the first year, many studies suggest that in perception and
production, in phonetics and phonology, their knowledge continues to
develop throughout childhood (e.g., Hazan \& Barrett, 2000). One common
task in this line of research is nonword repetition (NWR). In NWR
studies, participants are presented auditorily with an item that is
phonologically legal but lexically meaningless in the language children
are learning. The child should immediately try to say it back without
changing anything. NWR scores are thought to reflect long-term
phonological knowledge (which allows the child to perceive the item
precisely even though it is not a real word they have encountered
before) as well as online phonological working memory (to encode the
item in the interval between hearing it and saying it back) and flexible
production patterns (to produce the item precisely even though it had
never been pronounced before).

NWR has been used to seek answers to a variety of theoretical questions,
including what the links between phonology, working memory, and the
lexicon are (Bowey, 2001), and to assess the extent to which apparent
phonological constraints found in the lexicon affect online production
(Gallagher, 2014). It is also frequently used for applications, notably
as a diagnostic for language delays and disorders (Estes, Evans, \&
Else-Quest, 2007). Since non-words can be generated in any language, it
has attracted the attention of researchers working in multilingual and
linguistically diverse environments, particularly in Europe (Action,
2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has been seldom
used outside of Europe and North America (with exceptions including
Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz
(2020)).

In this study, we report on NWR results among children learning Yêly
Dnyé, an isolate spoken in Rossel Island, PNG, with an unusually dense
phonological inventory. The present report uses these unique data to
contribute to several lines of research. First, we made sure that some
of the non-word items contained typologically rare and/or challenging
sounds, so that we can contribute to the study of whether rare sounds
are disadvantaged in perception and/or production, both in terms of
overall repetition scores and patterns of mispronunciation. Second, we
look at the effects of item length, since previous NWR research has
uncovered variable effects. Third, we contribute to the basic NWR
literature by adding a data point for children learning a rarely-studied
language and culture, and further investigating whether there are
structured sources that account for individual variation.

\subsubsection{Yélî Dnye phonological
overview}\label{yuxe9luxee-dnye-phonological-overview}

Yélî Dnye is an isolate language (presumed Papuan) spoken by
approximately 7,000 people residing on Rossel Island, which is found at
the far end of the Louisiade Archipelago in Milne Bay Province, Papua
New Guinea. The Yélî sound system, much like its baroque grammatical
system (Levinson, in preparation), is unlike any other in the region.

With only four primary places of articulation (bilabial, alveolar,
post-alveolar, and velar) and no voicing contrasts, the phonological
inventory is remarkably packed with acoustically similar segments. The
core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and
doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal
equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
substantial portion of these able to be contrastively pre-nasalized or
nasally released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/,
/kŋ/, /ṭpṇm/, /kpŋm/). A large number of this combinatorial set can
further be contrastively labialized or palatalized on release, or both
(e.g., /pʲ/, /pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (in preparation)
for details). The consonantal inventory also includes a number of
non-nasal continuants (/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in
Yélî Dnye may be oral or nasal, short or long. The 10 oral vowel
qualities, which span four levels of vowel height, (/i/, /ɯ/, /u/, /e/,
/o/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/) can be produced as short or long vowels,
with 7 of these able to appear as short and long nasal vowels as well
(/i/, /u/, /ə/, /ɛ/, /ɔ/, /æ/, /ɑ/).

In total, Yélî Dnye uses 90 distinctive segments (93 when including
consonants that are extremely rarely used), far outstripping the
phonemic inventory size of other documented Papuan languages (Levinson
(in preparation); Foley 1986) and includes at least two contrasts not
yet been documented elsewhere {[}labial-coronal double-articulations
with dental vs.~post-alveolar coronal placement in both oral and nasal
stops, Maddieson \& Levinson 2001; Ladefoged and Maddieson 1996{]}.

Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
that, and with tri-and-above syllabic words appearing least frequently
(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the
current Yélî Dnye dictionary, Levinson, in preparation). The vast
majority of syllables use a CV format. A small portion of the lexicon
features words with a final closed syllable, limited to codas of -/m/,
-/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
process is used frequently with English loan words that have a coda
(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
these are limited to /æ/-. Finally, the lexicon features a handful of
single-vowel grammatical morphemes (see Levinson (in preparation) for
details).

While all consonants can appear in the first syllable of the word, the
second syllable onset consonant in bisyllabic roots is almost
exclusively drawn from a subset of the simplex consonants (in order of
descending frequency: /m/, /ṭ/, /n/, /p/, /l/, /w/, /j/, /pʷ/, /ŋ/, /k/,
/t/, /ɣ/, /mʷ/). In monosyllabic stems, oral vowels (short: 43\% and
long: 31\%) are more common than nasal ones (short: 4\% and long: 22\%),
with the first syllable of bisyllabic stems following a similar pattern.
Most vowels in the second syllable of bisyllabic stems are also short
and oral, with a potential preference for rounding harmony between the
first and second vowels. Stress typically falls on the first syllable of
bisyllabic words, on the first and third syllables of tetrasyllabic
words, and on the first or second syllable of trisyllabic words,
depending on its vocalic features (see Levinson (in preparation) for
details).

@@

The current study also contributes to increasing the cultural diversity
of the populations for which NWR data have been collected. As mentioned
previously, this method has been used almost exclusively in urban,
industrialized populations, with only one exception to our knowledge
(Cristia et al., 2020). We therefore provide some background information
tol help understand some aspects of our study, including the estimation
of children's ages and the relationship between NWR scores and
children's demographic characteristics.

Most speakers of the language grow up speaking it monolingually until
they begin attending school around the age of 7; school instruction is
in English. While monolingual Yélî Dnye upbringing is common,
multilingual families are not unusual, particularly in the region around
the Catholic Mission---the same region in which the current data were
collected---, where there is a higher incidence of married-in mothers
from other islands (Brown \& Casillas, in press). Children in these
multilingual families grow up speaking Yélî Dnye and English, Tok Pisin,
and/or some other language(s) from the region.

@@

Laying 250 nautical miles off the coast of mainland PNG and surrounded
by barrier reef, transport to and from the island is both infrequent and
irregular. There is no electricity system (we use solar panels) or motor
vehicles on the island, save dinghys with outboard motors that are
primarily reserved for medical transport, when there is deisel
available. There is a medical radio connection at the health clinic (at
the Catholic mission) via which news and messages are often passed
along. There is also a cellular tower on the island, though it is
typically not functioning. As such, access to news and outside
resources, particularly international connections and larger data
transfers, is typically unavailable. Our data collection is therefore
limited to the duration of the researchers' on-island visits.

Despite restricted outside contact, formal education is a priority for
many Yélî families. A recent study surveying more than 40 parents of
young children in this region of the island found that nearly all
parents completed 6 or more years of education, with 65\% of mothers and
49\% of fathers completing the full range of locally available education
(\textasciitilde{}9 years), with around half of those parents having
pursued an additional 2+ years of education on other islands in the
region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While
there are a handful of local schools around the island, these may be
well out of walking distance for many children (i.e., more than 1 hour
on foot or by canoe each day), and so it is very common for households
to billet school children during the weekdays for months at a time (or
for whole months if the child's home is very far). Combined with
practices of collective child guardianship within their close-knit
hamlets, the practical outcome for researchers is that adult consent can
often come from a combination of aunts, uncles, adult cousins, and
grandparents standing in for the child's parents---to this child assent
is also culturally pertinent, as child independence is encouraged and
respected from toddlerhood onward (Brown \& Casillas, in
press).\footnote{Note that in a few cases this does mean that precise
  details such as child date of birth, parent age in years, and parent
  education was only able to be approximated as the parents were
  residing on another part of the island.} Finally, as mentioned, the
language of schooling in this region is English, and so the majority of
teens and adults on the island speak some English---many speak it very
fluently.

While work on Yélî language development is growing (e.g., Brown, 2011,
2014; Brown \& Casillas, in press; Casillas et al., 2020; Liszkowski,
Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's
linguistic development is quite limited, and research into their
phonological development in particular has only just begun {[}e.g.,
Bram's thesis; MC's CFH data{]}. What we do know about the early
language environment is that, while child-directed speech is relatively
infrequent throughout the day (Casillas et al., 2020), shared caregiving
practices and a near-universal fondness for social interaction with
young children ensure that speech directed to children comes from all
types of speakers: women, men, and other children (Bunce et al., under
review; Casillas et al., 2020). While speech from women predominates in
young children's speech environments on Rossel, as it does elsewhere
(Bergelson et al., 2019; Bunce et al., under review), there is a
significant and marked increase in child-directed speech from other
children as infants get older (Bunce et al., under review; Casillas et
al., 2020). This increase in child-directed speech from other children
is attributed to the fact that, starting around age 2, children often
spend large swaths of the day playing, swimming, and foraging for fruit,
nuts, and shellfish in large (\textasciitilde{}10 people) independent
and mixed-age child play groups (Brown \& Casillas, in press; Casillas
et al., 2020).

\subsubsection{Brief review of NWR for our
purposes}\label{brief-review-of-nwr-for-our-purposes}

The basic structure of NWR involves the participant hearing a non-word,
like ``bilik'', which they need to repeat back without changing anything
that is contrastive in the language. For instance, in English, a
response of ``bilig'' or ``pilik'' will be scored as incorrect; a
response ``biilik'', where the vowel is lengthened without change of
quality would be scored as correct, because English does not have a pure
length contrast.

Aside from this central basic structure, there is some variation in the
presentation procedure and structure of items found in previous NWR
work. For example, items are often presented orally by the experimenter
(Torrington Eaton, Newman, Ratner, \& Rowe, 2015), although an
increasing number of studies have turned to playing back the stimuli in
order to have greater control of the stability of the presentation
(Brandeker \& Thordardottir, 2015). Additionally, while some studies
have used 10-15 non-words, others have employed up to 46 unique items
(Piazzalunga, Previtali, Pozzoli, Scarponi, \& Schindler, 2019). Often,
authors modulate structural complexity, typically measured in terms of
item length (measured in number of syllables) and/or syllable structure
(open as opposed to closed syllables, Gallon, Harris, \& Van der Lely,
2007).

Previous work seems to avoid difficult sounds, but we felt this was
important to represent Yélî Dnye, so we also varied this factor. We
designed a relatively large number of items but, aware that this may
render the task longer and more tiresome, we split some of the items
across children. This allowed us to get NWR scores for more items.

Naturally, designing the task in this way may render the study of
individual variation within the population more difficult because
different children are exposed to different items. However, a review of
previous work on individual variation suggested to us that effects of
individual differences are relatively small, and would not be detectable
with the sample size that we could collect in a given visit.

That said, we contribute to the literature by also reporting descriptive
analyses of individual variation that could potentially be integrated in
meta-analytic efforts. Based on previous work, we looked at potential
improvements with age (Farmani et al., 2018; Kalnak, Peyrard-Janvid,
Forssberg, \& Sahlén, 2014; Vance, Stackhouse, \& Wells, 2005), and
potential negative effects of bilingual exposure (Brandeker \&
Thordardottir, 2015; Meir \& Armon-Lotem, 2017; Meir et al., 2016).
Previous work typically finds no significant differences as a function
of maternal education (e.g., Farmani et al., 2018; Kalnak et al., 2014;
Meir \& Armon-Lotem, 2017) or child gender (Chiat \& Roy, 2007).
Although past research has not often investigated potential effects of
birth order on NWR, there is a sizable literature on these effects in
other language tasks (Havron et al., 2019), and therefore we report on
those too.

@@

\begin{itemize}
\tightlist
\item
  Does the frequency of sounds across languages predict NWR? Are rarer
  sounds more often substituted by commoner sounds?
\item
  How does score change as a function of item length in number of
  syllables?
\item
  Is individual variation attributable to child age, sex, birth order,
  monolingual status, and/or parental education? 
\end{itemize}

In view of the hypothesis-driven nature of this work, we had considered
boosting the interpretational value of this evidence by announcing our
analysis plans prior to conducting them. However, we realized that even
pre-registering an analysis would be equivocal because we do not have
enough power to look at all relationships of interest, and often to
detect any of the known effects given their variability across studies.
To illustrate this, we portray studies in which children's NWR scores
were gathered between 4 and 12 years of age, and reported separately for
items that are relatively short (1-2 syllables) versus longer items (3-4
syllables) in Figure 1. Notice that the effect of stimulus length is
minuscule among Italian children (Piazzalunga et al., 2019), but
considerable among Tsimane' children (Cristia et al., 2020), where a
drop of 40 percentage points is observed at all ages. A similar
difference in NWR scores for short versus long items was observed among
Arabic children (Jaber-Awida, 2018). Even the effect of age is unstable
in this sample. Whereas it is quite clear that children's NWR scores
increase in the Italian data, age effects are less stable among Tsimane'
children. Therefore, all analyses here are descriptive and should be
considered exploratory.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in
length between 1 and 4 syllables, often additionally varying syllable
complexity and/or cluster presence and complexity, always meeting the
condition that they do not mean anything in the target language (e.g.,
Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We kept the
same variation in item length and the non-meaningfulness requirement,
but we did not vary syllable complexity and clusters because these are
vanishingly rare in Yély Dnye. We also increased the number of items an
individual child would be tested on, so that a child would get up to 23
items to repeat (note that up to other work has also used 24-30 items:
Jaber-Awida, 2018; Kalnak et al., 2014), and we created more items and
distributed them across children, so as to increase the coverage, and be
able to study more items.

A first list of candidate items was generated in 2018 by selecting
simple consonants (``p'', ``t'', ``d'', ``k'', ``m'', ``n'', ``w'',
``y'') and vowels (``i'', ``o'',``u'', ``a'',``e'') that were combined
into consonant-vowel syllables, further sampling the space of 1- to
4-syllable sequences. These candidates were automatically checked
against Levinson's 2015 dictionary and removed from consideration if
they appeared in the dictionary. The second author presented them orally
to three local research assistants, who were asked to repeat them and
further say whether they were real words. Any item for which two or more
of the assistants reported them having a meaning or some form of
association was excluded.

>>>>

\abstract{
In nonword repetition (NWR) studies, participants are presented
auditorily with an item that is phonologically legal but lexically
meaningless in the local language. NWR scores are thought to reflect
long-term phonological knowledge as well as online phonological working
memory and flexible production patterns. In this study, we report on NWR
results among children learning Yélî Dnye, an isolate spoken on Rossel
Island in Papua New Guinea, with an unusually dense phonological
inventory. This study contributes to three lines of research. First, we
document that non-word items containing typologically rare sounds are
repeated without changes less often that non-words containing more
common sounds. Second, we document rather weak effects of item length,
contributing to other research suggesting that length effects may be
language-specific. Third, we do not find strong individual variation
effects in this population, contrary to previous results documenting
strong age-related effects. Together, these data provide a unique view
of online phonological processing in a seldom-studied language, and
contribute to both typological and language acquisition research.

@@

\begin{itemize}
\tightlist
\item
  check again if all fields are explained
\item
  set up OSF (or revise if already there)
\item
  implement decisions on prop phones \& patterns nwr mp
\item
  revise bullet points for discussion
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  double check my frequency entries in segments.xlsx using the search
  function in \url{https://phoible.org/parameters}; take a look at
  \url{http://phoible.github.io/conventions/} in case you see something
  about double articulation (I didn't find tp or lBj) 
\item
  draft discussion
\end{itemize}

@@

Children's perception and production of phonetic and phonological units
continues developing well beyond the first year of life, even extending
into middle childhood (e.g., Hazan \& Barrett, 2000). Much of the
evidence for later phonological development comes through the use of
nonword repetition (NWR) tasks. In a NWR task, participants hear a short
word-like form that is phonologically legal but lexically meaningless in
the language(s) they are learning. After hearing this non-word, the
participant's task is to try to immediately and precisely repeat it. NWR
scores are thought to reflect long-term phonological knowledge (to
perceive the item precisely despite not having heard it before) as well
as online phonological working memory (to encode the item in the
interval between hearing it and saying it back) and flexible production
patterns (to produce the item precisely despite not having pronounced it
before).

NWR has been used to seek answers to a variety of theoretical questions,
including what the links are between phonology, working memory, and the
lexicon (Bowey, 2001), and how extensively apparent phonological
constraints found in the lexicon affect online production (Gallagher,
2014). NWR is also frequently used in applied contexts, notably as a
diagnostic tool for language delays and disorders (Estes, Evans, \&
Else-Quest, 2007). Since non-words can be generated in any language, it
has attracted the attention of researchers working in multilingual and
linguistically diverse environments, particularly in Europe (Action,
2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has seldom been
used outside of Europe and North America (with exceptions including
Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz
(2020)). Further, the languages it has been used with have historically
had small to moderate phonological inventories (Maddieson, 2005).

In the present study, we use NWR to investigate the phonological
development of children learning Yélî Dnye, an isolate language spoken
in Papua New Guinea (PNG) that has a large and unusually dense
phonological inventory. The study was designed to contribute to multiple
aspects of our understanding of phonological development: First, we
included a subset of non-word items with typologically rare and/or
challenging sounds to ask whether these rare sounds are disadvantaged in
perception and/or production, both in terms of overall repetition scores
and patterns of mispronunciation. Second, we varied the length (in
syllables) of non-words to investigate the impact of word length on
repetition and mispronunciation; previous NWR research has shown
variable effects of length between populations. Third, these data give a
first impression of how this task is responded to in a rarely-studied
language and culture; we add to this by further investigating whether
there are structured sources that account for individual variation in
our sample.

Before jumping into the details of our study design we first give an
overview of Yélî Dnye phonology as well as a brief ethnographic review
of the developmental environment on Rossel Island. Since NWR has been
almost exclusively used in urban, industrialized populations (see
Cristia et al., 2020 for an exception), we provide this additional
ethographic information to contextualize the adaptations we have made in
running the task and in gathering age and other demographic information,
compared to what is typical in urban, industrialized settings.

\subsubsection{Yélî Dnye phonology}\label{yuxe9luxee-dnye-phonology}

Yélî Dnye is an isolate language (presumed Papuan) spoken by
approximately 7,000 people residing on Rossel Island, an island found at
the far end of the Louisiade Archipelago in Milne Bay Province, Papua
New Guinea. The Yélî sound system, much like its baroque grammatical
system (Levinson, accepted), is unlike any other in the region.

With only four primary places of articulation (bilabial, alveolar,
post-alveolar, and velar) and no voicing contrasts, the phonological
inventory is remarkably packed with acoustically similar segments. The
core oral stop set includes both singleton (/p/, /t/, /ṭ/, and /k/) and
doubly-articulated (/tp/, /ṭp/, /kp/) segments, with full nasal
equivalents (/m/, /n/, /ṇ/, /ŋ/, /nm/, /ṇm/, /ŋm/), and with a
substantial portion of them contrastively pre-nasalized or nasally
released (/mp/, /nt/, /ṇṭ/, /ŋk/, /nmtp/, /ṇmṭp/, /ŋmkp/, /ṭṇ/, /kŋ/,
/ṭpṇm/, /kpŋm/). A large number of this combinatorial set can further be
contrastively labialized, palatalized on release, or both (e.g., /pʲ/,
/pʷ/, /pʲʷ/; /tpʲ/; /ṇmḍbʲ/; see Levinson (accepted) for
details).\footnote{We use Levinson's (accepted) under-dot notation
  (e.g., /ṭ/) to indicate the post-alveolar place of articulation; these
  stops are, articulatorily, somewhat variable in place, with at least
  some tokens produced fully sub-apically. In approximating
  cross-linguistic segment frequency below we use the corresponding
  retroflex for each stop segment (e.g., /ʈ/, /ʈp/, /ɳ/).} The
consonantal inventory also includes a number of non-nasal continuants
(/w/, /j/, /ɣ/, /l/, /βʲ/, /lʲ/, /lβʲ/). Vowels in Yélî Dnye may be oral
or nasal, short or long. The 10 oral vowel qualities, which span four
levels of vowel height, (/i/, /ɯ/, /u/, /e/, /o/, /ə/, /ɛ/, /ɔ/, /æ/,
/ɑ/) can be produced as short and long vowels, with 7 of these able to
appear as short and long nasal vowels as well /ĩ/, /ũ/, /ə̃/, /ɛ̃/, /ɔ̃/,
/æ̃/, /ɑ̃/).

In total, Yélî Dnye uses 90 distinctive segments (not including an
additional 3 rarely used consonants), far outstripping the phonemic
inventory size of other documented Papuan languages (Foley, 1986;
Levinson, accepted) and includes at least two contrasts not yet been
documented elsewhere (labial-coronal double-articulations with dental
vs.~post-alveolar coronal placement for both oral and nasal stops,
Ladefoged \& Maddieson, 1996; Maddieson \& Levinson, in preparation).

Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
that, and with tri-and-above syllabic words appearing least frequently
(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most
recent Yélî Dnye dictionary at time of writing, Levinson, accepted). The
vast majority of syllables use a CV format. A small portion of the
lexicon features words with a final closed syllable, limited to codas of
-/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
process is used frequently with English loan words that have a coda
(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
these are limited to /æ/-. Finally, the lexicon features a handful of
single-vowel grammatical morphemes (see Levinson (accepted) for
details).

@@

Most speakers of Yélî Dnye grow up speaking it monolingually until they
begin attending school around the age of 7; school instruction is in
English. While monolingual Yélî Dnye upbringing is common, multilingual
families are not unusual, particularly in the region around the Catholic
Mission---the same region in which the current data were
collected---where there is a higher incidence of married-in mothers from
other islands (Brown \& Casillas, in press). Children in these
multilingual families grow up speaking Yélî Dnye plus English, Tok
Pisin, and/or another language(s) from the region.

@@

Laying 250 nautical miles off the coast of mainland PNG and surrounded
by a barrier reef, transport to and from the island is both infrequent
and irregular. There is no electricity system (we use solar panels) or
motor vehicles on the island, save dinghys with outboard motors that,
even when there is deisel available, are primarily reserved for medical
transport. There is a medical radio connection at the health clinic,
located at the Catholic Mission, via which news and messages are often
passed to and from mainland PNG. There is also a cellular tower on the
island, though it is often inoperative, usually for multiple months or
years at a time. As such, access to news and outside connections,
including international phone calls and digital exchanges that require
significant data transfer, is typically unavailable. Our data collection
is therefore limited to the duration of the researchers' on-island
visits.

Despite restricted outside contact, formal education is a priority for
many Yélî families. A recent study surveying more than 40 parents of
young children in this region of the island found that nearly all
parents completed 6 or more years of education, with 65\% of mothers and
49\% of fathers completing the full range of locally available education
(\textasciitilde{}9 years), with around half of those parents having
pursued an additional 2+ years of education on other islands in the
region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While
there are a handful of local schools around the island, these may be
well out of walking distance for many children (i.e., more than 1 hour
on foot or by canoe each day). It is very common for households situated
close to schools to billet their school-aged relatives during the
weekdays (sometimes the weekends too, if the child's home is very far)
for long segements of each school year. Combined with practices of
collective child guardianship within each close-knit hamlet, the
practical consequence of this billeting for researchers is that adult
consent can often come from a combination of aunts, uncles, adult
cousins, and grandparents standing in for the child's biological
parents. On top of this, child assent is culturally pertinent, as child
independence is encouraged and respected from toddlerhood onward (Brown
\& Casillas, in press).\footnote{For these reasons, we were unable to
  record precise details for a few children, including child date of
  birth, parent age in years, and parent education.} Finally, as
mentioned, the language of schooling in this region is English, and so
the majority of teens and adults on the island speak some English---many
speak it very fluently.

While work on Yélî language development is growing (e.g., Brown, 2011,
2014; Brown \& Casillas, in press; Casillas et al., 2020; Liszkowski,
Brown, Callaghan, Takada, \& de Vos, 2012), our knowledge of children's
linguistic development is quite limited, and research into their
phonological development in particular has only just begun (e.g., Peute,
Fikkert, \& Casillas, In preparation). What we do know about the early
language environment is that, while direct speech to children under 3;0
is relatively infrequent throughout the day (Casillas et al., 2020),
shared caregiving practices and a near-universal fondness for social
interaction with young children ensure that speech directed to children
comes from all types of speakers: women, men, and other children (Bunce
et al., under review; Casillas et al., 2020). While speech from women
predominates in young children's speech environments on Rossel, as it
does elsewhere (Bergelson et al., 2019; Bunce et al., under review),
there is a significant and marked increase in child-directed speech from
other children as infants get older (Bunce et al., under review;
Casillas et al., 2020). This increase in child-directed speech from
other children is attributed to the fact that, starting around age 2,
children often spend large swaths of the day playing, swimming, and
foraging for fruit, nuts, and shellfish in large (\textasciitilde{}10
members) independent and mixed-age child play groups (Brown \& Casillas,
in press; Casillas et al., 2020).

\subsubsection{NWR design and analysis
adaptations}\label{nwr-design-and-analysis-adaptations}

In a basic NWR task, the participant listens to a production of a
word-like form, such as /bilik/, and then repeats back what they heard
without changing any phonological feature that is contrastive in the
language. For instance, in English, a response of {[}bilig{]} or
{[}pilik{]} would be scored as incorrect; a response {[}biilik{]}, where
the vowel is lengthened without change of quality would be scored as
correct, because English does not have a general feature of contrastive
vowel length. There is some variation in how past NWR studies have
designed the presentation procedure and structure of items. For example,
while items are often presented orally by the experimenter (Torrington
Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies
have turned instead to playing back pre-recorded stimuli in order to
increase control in stimulus presentation (Brandeker \& Thordardottir,
2015). Additionally, while some studies have used 10-15 non-words,
others have employed up to 46 unique items (Piazzalunga, Previtali,
Pozzoli, Scarponi, \& Schindler, 2019). Authors also often modulate
structural complexity, typically measured in terms of item length
(measured in number of syllables) and/or syllable structure (open as
opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).

Previous work typically steers clear of articulartorily and/or
acoustically challenging sounds, but we included some in our experiment
to more adequately represent Yélî Dnye's phonology. We ultimately used a
relatively large number of items that would enable us to explor both
variation in structural complexity and in more vs.~less challenging
sounds. However, aware that this large item inventory might render the
task longer and more tiresome, we split items across children (see
below). Naturally, designing the task in this way may render the study
of individual variation within the population more difficult because
different children are exposed to different items. However, a review of
previous work on individual variation suggested to us that effects of
individual differences in NWR are relatively small, and would not be
detectable with the sample size that we could collect during our short
visit. That said, we contribute to the literature by also reporting
descriptive analyses of individual variation that could potentially be
integrated in meta-analytic efforts.

Based on previous work, we looked at potential changes with age (Farmani
et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance,
Stackhouse, \& Wells, 2005), and differences relating to bilingual
language exposure (Brandeker \& Thordardottir, 2015; Meir \&
Armon-Lotem, 2017; Meir et al., 2016). Previous work typically finds no
significant differences as a function of maternal education (e.g.,
Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or
child gender (Chiat \& Roy, 2007). Although past research has not often
investigated potential effects of birth order on NWR, there is a sizable
literature on these effects in other language tasks (Havron et al.,
2019), and therefore we report on these too.

@@

\begin{itemize}
\tightlist
\item
  Does the cross-linguistic frequency of sounds predict NWR scores? Are
  rarer sounds more often substituted by commoner sounds?
\item
  How does score change as a function of item length in number of
  syllables?
\item
  Is individual variation attributable to child age, sex, birth order,
  monolingual status, and/or parental education?
\end{itemize}

In view of the hypothesis-driven nature of this work, we had considered
boosting the interpretational value of this evidence by announcing our
analysis plans prior to conducting them. However, we realized that even
pre-registering an analysis would be equivocal because we would not have
enough power to look at all relationships of interest, in many cases
possibly not enough to detect any of the known effects, given their
variability across studies. To illustrate this issue, we portray in
Figure 1 studies in which children's NWR scores were gathered between 4
and 12 years of age, and reported separately for items that are
relatively short (1-2 syllables) versus longer items (3-4 syllables).
Notice that the effect of stimulus length is minuscule among Italian
children (Piazzalunga et al., 2019), but considerable among Tsimane'
children (Cristia et al., 2020), where a drop of 40 percentage points is
observed at all ages. A similar large difference in NWR scores for short
versus long items was observed among Arabic children (Jaber-Awida,
2018). Even the effect of age is unstable in this sample of studies.
Whereas it is quite clear that children's NWR scores increase in the
Italian data, age effects are less stable among Tsimane' children.
Therefore, all analyses in the present study are descriptive and should
be considered exploratory.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in
length between 1 and 4 syllables, often additionally varying syllable
complexity and/or cluster presence and complexity, and always meeting
the condition that they do not mean anything in the target language
(e.g., Balladares, Marshall, \& Griffiths, 2016; Wilsenach, 2013). We
kept the same variation in item length and requirement for not being
meaningful in the language, but we did not vary syllable complexity or
clusters because these are vanishingly rare in Yélî Dnye. We also
increased the number of items an individual child would be tested on,
such that a child would get up to 23 items to repeat (other work has
also used up to 24-30 items: Jaber-Awida, 2018; Kalnak et al., 2014),
with the entire test inventory of 44 items distributed across children.

A first list of candidate items was generated during a trip to the
island in 2018 by selecting simple consonants (/p/, /t/, /ṭ/, /k/, /m/,
/n/, /w/, /y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them
into consonant-vowel syllables, then sampling the space of 1- to
4-syllable sequences. These candidates were automatically removed from
consideration if they appeared in Levinson's (2015) dictionary. The
second author presented them orally to three local research assistants,
all native speakers of Yélî Dnye, who were repeated each form, as they
would in an NWR task, and additinally let the experimenter know if the
item was in fact a word or phrase in Yélî Dnye. Any item reported to
have a meaning or a strong association with another word form or meaning
was excluded.
